---
title: "Count-versus-length study design"
bibliography: references.bib
toc: true
format:
  html:
    code-tools: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: length-design-setup
read_design <- function(name) readr::read_csv(here::here("outputs", "tables", paste0("recruitment-length-", name, ".csv")), show_col_types = FALSE)
flow <- read_design("opportunity-flow")
precision <- read_design("paired-precision")
recall <- read_design("recall-precision")
example_precision <- precision[precision$half_width == .01 & precision$variance_inflation == 1.5 & precision$retention == .8, ]
```

## Research objective

This optional follow-up design builds on the completed Michigan
[recruitment study](recruitment.qmd). It specifies the additional measurements
and evaluation sample needed to assess seedling length. Regional acquisition
and external validation are future work.

Determine whether the length distribution of established sugar-maple seedlings
improves predictions of recorded sapling entry beyond ordinary seedling counts
on the same sampled area. The immediate scientific product is a paired
measurement comparison with a credible assessment sample. A survey-ranking
tool, a treatment prescription, and a forecast of canopy replacement are
different products requiring additional evidence.

The Michigan [measurement audit](ri-audit.qmd) supplies local feasibility
evidence: 90
baseline intervals on 87 physical plots contain 14 recruit-bearing conditions.
The 71 intervals with ordinary baseline seedling detections contain all 14
events. Taller-seedling presence is associated with subsequent entry, but its
added value after accounting for abundance has not been estimated. The current
later-year partition supplies only four assessment events.

Regional research already compared ordinary counts with six RI length classes
and found that taller seedlings were important predictors. The proposed
one-term comparison is therefore a parsimonious regional assessment, not the
discovery that height matters or an exact replication of the published boosted
tree models. [@harris2022]

## Local sampling opportunities

The existing Michigan database was screened across baseline RI visits, rather
than only the current evaluation cohort. Baselines require a sampled national-design
plot, accessible northern-hardwood forest type, positive microplot area,
and a live sugar maple at least five inches DBH. The active forest codes are
801, 802, 805, and 809. Successor plot records, measurement dates, and ordinary
baseline seedling-record presence then define the following opportunity screen.

```{r}
#| label: tbl-length-opportunities
#| tbl-cap: "RI record and visit-date opportunities in the existing Michigan snapshot. These are not validated recruitment intervals or event counts. The final row is a separate branch."
flow |> dplyr::transmute(Screen = .data$screen,
  `Condition visits` = .data$baseline_condition_visits,
  `Physical plots` = .data$physical_plots, Counties = .data$counties) |> knitr::kable()
```

There are 234 qualifying baseline condition visits on 116 physical plots.
Of those, 128 have exactly one linked successor; 114 also have a sampled
national-design successor and a recorded interval of 4.5-8.5 years. Requiring
a positive ordinary seedling record leaves 88 opportunities on 79 plots.

Those 88 comprise 71 already included in the current RI cohort, ten additional
condition visits on plots represented in the current recruitment cohort, and
seven visits on six plots outside that cohort. Physical-plot counts cannot be
added across these groups because a plot can supply several historical visits.
This is a small potential extension, not a new independent evaluation sample.

The screen does not establish RI microplot status, source/count validity,
condition continuity, shared-area eligibility, or sapling-entry events for the
additional visits. It queries no new recruitment outcomes. The wider historical
Michigan source is not an untouched validation sample; unused rows or predictors
do not make it one. The 106 baseline condition
visits without a linked successor are unresolved opportunities in this snapshot,
not guaranteed future observations or a scheduled data release.

As of September 9, 2026, the official Michigan update history lists addition
of the 2025 evaluation on April 27, followed by soil and down-wood table updates.
It does not identify a newer recruitment inventory cycle. This is release-note
evidence, not a bytewise comparison of a freshly downloaded database; refreshing
the file cannot be assumed to create an independent cohort. [@fiaupdates2026mi]

## Primary population and endpoint

The primary comparison is among conditions with ordinary baseline sugar-maple
seedlings, not the easier mixture of detected and undetected seedlings. Retain
the northern-hardwood/adult-maple scope and verified common sampling opportunity
from the recruitment study. A regional extension must use harmonized definitions
and report each geography separately; a result outside Michigan is a transport
assessment, not independent validation of Michigan-specific performance.

The outcome remains at least one new live sugar-maple sapling recorded at the
next visit: species 318, live status, diameter at least one and below five
inches, reconciliation code 1, no predecessor TREE identifier, and the audited
common microplot frame. It excludes ambiguous procedure changes and missed-tree
tallies. It is not an individual seedling conversion rate: intervening entrants
that die before remeasurement are not observed.

Apply the existing baseline RI frame and count checks before assigning any tally
zero. Preserve invalid or inaccessible observations as unavailable. Historical
manual access remains qualified, including the 2016 archived draft and the
unrecovered 2012-2015 annual supplements. A broader cohort needs its own
year-specific review rather than inheriting blanket approval from the 90 cases.

## Fixed comparison

Let N denote the ordinary baseline maple count, H the RI count in classes 5-6
(at least five feet long), A the sampled microplot coverage, and T the elapsed
interval. The proposed length feature is Q = H/N. It is undefined when N is
zero; those conditions are outside the primary comparison, not assigned Q = 0.

| Component | Count model | Count-plus-length model |
|---|---|---|
| Abundance | log(1 + N) | log(1 + N) |
| Sampling opportunity | A | A |
| Recorded timing | T | T |
| Length structure | None | Q = H/N |
| Model family | Ridge logistic regression | Same |
| Penalty | 1 on standardized slopes; intercept unpenalized | Same |
| Non-intercept terms | 3 | 4 |

Both models use identical cases and development-only standardization. Do not
add beech, browse score, six separate classes, interactions, alternative cutoffs,
or a favorable model family after inspecting evaluation results. This choice
limits the question to one candidate length representation; a negative result
would not prove that no length-based method could help.

A required, predeclared sensitivity replaces the single log-count slope with
a three-degree-of-freedom natural spline of log(1 + N) in both models, while
retaining A, T, and the same single Q addition. Set knots and boundary knots
from development data only, standardize the resulting basis there, and use
the same fixed ridge penalty. This gives five versus six slopes and must be
included in development-support planning. Report it alongside the primary
linear-log-count comparison, not as a replacement chosen after seeing results.
If the length gain disappears against flexible counts, limit the conclusion to
the original model forms; do not claim established added measurement value.

Ordinary counts must be reconciled with RI classes 3-6 on the same sampled area.
Unresolved differences stop the comparison before fitting: silently retaining
only agreements could create a selected sample. RI seedlings below one foot
are excluded from N and Q. Their potential information is a separate question
and is not mixed into this one-term length comparison.

For each physical plot, choose the latest completed eligible visit interval
within the declared development or assessment period using dates and protocol
eligibility, not outcome values. Retain eligible conditions from that interval.
Keep every visit and condition of a physical plot on the same side of the
development/assessment boundary. Give each plot total training weight one,
divided across its included conditions. Freeze this interval-selection and
weighting rule before any new outcome comparison.

## Timing and the claim being tested

The primary estimand is a retrospective next-visit probability comparison
conditional on the realized interval within 4.5-8.5 years. Realized T is shared
by both models, so the comparison can test added length information after
adjusting for observation duration. However, the actual next-visit date is not
known at baseline. This estimand is not an operational forecast made at baseline.

A future forecast must instead specify a genuinely planned horizon and collect
or identify outcomes at that horizon. Merely setting T = 7 when scoring existing
variable-duration outcomes does not validate seven-year risks. There is no
observed exact crossing date from which to reconstruct an exact seven-year
recruitment event. The current study should retain the narrower conditional
claim until suitable timing evidence exists.

## Evaluation and meaningful improvement

The primary metric is paired Brier gain: count-model squared probability error
minus length-model squared error. Average the condition-level differences
within each physical plot, then average across plots:

$$
D_j = \frac{1}{m_j}\sum_{i=1}^{m_j}
\left[(y_{ij}-p_{0,ij})^2-(y_{ij}-p_{1,ij})^2\right],
\qquad \Delta = \frac{1}{J}\sum_{j=1}^{J}D_j.
$$

Positive gain favors length. Equal plot weighting prevents plots with more
conditions from automatically dominating the evaluation. It differs from a
condition-average estimator, which must not be substituted as the benchmark.
This is a selected-cohort comparison, not a statewide design-weighted
estimate of forest condition.

Evaluate the frozen models once on the reserved sample. Resample whole counties
for paired uncertainty, retaining all selected plot/condition records within
each county. Such evaluation intervals are conditional on the frozen training
fits; they do not include every source of model-development uncertainty. Report
geographic event support, calibration curves/intercept/slope, and discrimination
alongside Brier gain. With too few independent county clusters, a nominal
bootstrap interval is not a sufficient precision guarantee.

A lower confidence limit above zero supports a statistical improvement for
these models and this target. To claim a *meaningful* improvement, first declare
a minimum gain delta and require the lower limit to exceed it. An interval
spanning zero is inconclusive, not proof of equivalence. Neither an arbitrary
Brier threshold nor a small P value supplies field utility. The necessary
measurement costs and error consequences remain unspecified.

## Precision scenarios

External-assessment sample sizes depend on the target population, anticipated
predictions, calibration, and desired precision; an event-count rule alone
cannot establish adequacy. The published guidance concerns clinical models;
its design principles transfer, but its clinical thresholds and numerical
examples are not forestry sample-size recommendations. [@riley2024validation]

The following calculation is a transparent planning approximation for the
mean of frozen plot-level paired losses. With assumed SD s, variance inflation
V, and target 95% interval half-width h, use

$$
J_{\mathrm{complete}} = \left\lceil V(1.96s/h)^2\right\rceil.
$$

The implementation uses the exact normal quantile rather than rounded 1.96.
Candidate plot numbers are complete plots divided by an assumed retention
fraction, rounded upward. This scales expected yield; it does not guarantee
the required number will be retained, remove attrition bias, or certify power.

```{r}
#| label: tbl-length-precision
#| tbl-cap: "Hypothetical evaluation precision: half-width 0.01 Brier units, variance inflation 1.5, and 80% retention. No SD or inflation factor was estimated from a fitted RI model."
example_precision |> dplyr::transmute(`Assumed plot-loss SD` = .data$loss_difference_sd,
  `Complete evaluation plots` = .data$complete_evaluation_plots,
  `Candidates at 80% retention` = .data$candidate_plots_at_assumed_retention) |> knitr::kable()
```

For the middle SD assumption, this gives 577 complete evaluation plots and 722
candidates at 80% retention. The range is 145-2,305 complete plots across the
three SD assumptions. These are not estimates of the number available in any
state, not training sample sizes, and not proof of calibration precision. The
[full scenario table](../outputs/tables/recruitment-length-paired-precision.csv)
varies h, V, retention, and s. A smaller target gain typically demands greater
precision; h itself is not a minimum useful gain.

A separate fixed-rule illustration asks how many independent event units would
give a Wilson interval of a chosen width at an assumed recall. It uses the
score-interval formula documented by NIST. [@nistwilson]

```{r}
#| label: tbl-length-recall
#| tbl-cap: "Fixed-rule recall illustration. Widths are evaluated at an assumed proportion, not guaranteed or expected widths; these are event units, not total plots."
recall |> dplyr::transmute(`Assumed recall` = scales::percent(.data$assumed_recall),
  `Target half total width` = scales::percent(.data$target_half_total_width),
  `Independent event units` = .data$independent_event_units,
  `Plug-in interval` = paste0(scales::percent(.data$lower, accuracy = .1), " to ", scales::percent(.data$upper, accuracy = .1))) |> knitr::kable()
```

At assumed 80% recall, the half-total-width targets 0.10 and 0.05 require 60 and
245 independent event units, respectively. Wilson intervals are not symmetric
around the assumed proportion. These are not calculations for a paired gain in
recall, a data-selected capacity rule, or dependent conditions on the same plot.
They illustrate why four or fourteen events do not settle a forecasting claim.

Before recruitment or model fitting, replace the planning inputs with defensible
development and external-assessment ranges, assess training complexity and calibration
precision separately, and simulate the proposed cluster structure. Do not add
the planning sample counts together or automatically treat their maximum as a
fully justified study size: they address different quantities and assumptions.

## Next-stage decision

The local opportunity census does not reveal a sufficiently large independent
Michigan assessment cohort. More historical Michigan rows may inform development
or sensitivity analysis after full eligibility checks, but do not solve that
evaluation problem. Two distinct research routes remain:

| Route | Contribution and boundary |
|---|---|
| Michigan-only | Complete the measurement paper now; reserve genuinely new, adequately supported observations for a future conditional or planned-horizon assessment. A database refresh must be checked for new visits and plot overlap. |
| Regional extension | Screen Wisconsin and Minnesota as candidate Great Lakes sources under the same definitions, then fix development and geographically separate evaluation roles from protocol/metadata support before outcome-driven model selection. This changes the target geography and tests transport, not Michigan-specific validation. |

The recommended next action is a scoped regional **feasibility screen**, not
immediate model fitting. Neither state's usable RI sample or event support has
been established here. The official DataMart is the acquisition route; files
must be pinned with source dates and hashes, with measurement inventories
separated from evaluation-outcome access. [@fiadatamart]

Proceed to fitting only after source/geography approval, a documented eligible
cohort, a declared assessment set, separate development/evaluation precision
work, and a frozen analysis specification. If support is insufficient, retain
the descriptive study. Changing the population or adding predictors solely to
obtain a favorable result would not complete this research objective.

## Regional acquisition status

The regional feasibility screen covers Wisconsin and Minnesota under the same
scientific definitions. Acquisition was attempted, but neither ZIP could be
retrieved: the archive endpoints return an empty
response in repeated HTTP checks, and the DataMart landing page also fails in
the browser. **Regional sample size and validation feasibility remain unknown.**
This is an access blocker, not evidence of inadequate samples.

The separate live update-history service lists addition of the 2025 inventories
on May 4 for Wisconsin and May 11 for Minnesota, followed by soil-laboratory
updates on July 6. Those release notes do not establish usable RI/repeat-visit
support or the contents of a retrieved archive.
[Wisconsin history](https://apps.fs.usda.gov/fiadb-api/fiadb_update_history?statecd=55),
[Minnesota history](https://apps.fs.usda.gov/fiadb-api/fiadb_update_history?statecd=27).

The [acquisition audit](../outputs/audits/recruitment-regional-acquisition.csv)
records failed transfers with missing sizes and hashes, not zero observations.
`make regional-acquire` provides the bounded retry and source-pinning step;
it is intentionally separate from the Michigan pipeline. A successful transfer
must pass archive, state, table, integrity, and disk-space guards. The success
path is not yet exercised on the actual regional archives.

The [survey validation plan](../docs/survey-validation-plan.md) keeps regional
feasibility separate from the Michigan outcome analysis. Archive access remains
blocked, so no additional model, outcome comparison, or regional validation
result is supplied here.

## Reproducibility

`make length-plan` regenerates the metadata and hypothetical precision tables.
The helpers also implement the defined length fraction and plot-balanced paired
score for future use; they have not scored a new RI model. Record-level metadata
remain local and Git-ignored. No additional inventory was downloaded and no
model was fitted for this design.

The [survey validation plan](../docs/survey-validation-plan.md) and
[analysis decisions](../docs/analysis-decisions.md) retain the scientific and
operational boundaries.

## References

::: {#refs}
:::
