---
title: "Regeneration length classes and measurement feasibility"
description: "Michigan FIA regeneration-indicator sampling, count reconciliation, and support for a height-information comparison."
bibliography: references.bib
toc: true
format:
  html:
    code-tools: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: ri-setup
read_ri <- function(name) readr::read_csv(here::here("outputs", "tables", paste0("recruitment-ri-", name, ".csv")), show_col_types = FALSE)
flow <- read_ri("eligibility")
reconciliation <- read_ri("count-reconciliation")
zeros <- read_ri("zero-accounting")
height <- read_ri("height-support")
classes <- read_ri("length-classes")
split <- read_ri("split-support")
precision <- read_ri("precision-reference")
main <- flow[flow$screen == "Baseline RI available, including date screen", ]
all_height <- height[height$population == "All RI-available intervals", ]
tall <- all_height[all_height$group == "RI maple at least five feet present", ]
short <- all_height[all_height$group == "No RI maple at least five feet recorded", ]
pct <- function(x) scales::percent(x, accuracy = 0.1)
```

## Measurement findings

The Michigan regeneration-indicator (RI) records support a descriptive
measurement study, but not a strong new forecasting claim. The baseline
sampling and record checks retain `r main$conditions` condition intervals on
`r main$physical_plots` physical plots, with `r main$events` recruit-bearing
conditions. The audit examines whether seedling length measurements add
distinct information beyond the ordinary counts recorded on the same
sampled area.

Taller seedlings provide a plausible lead. Among `r tall$conditions` conditions
with RI maple seedlings at least five feet long, `r tall$events` had recorded
new saplings at follow-up; among `r short$conditions` without such seedlings,
`r short$events` did. This is a retrospective association in an existing sample.
It is not a held-out comparison with seedling abundance, proof of
individual survival, or a management threshold.

The information constraint remains decisive: the existing later-year split
would provide `r split$events[1]` development events and
`r split$events[2]` assessment events. No height-class model is fitted. The
current contribution is a reconciled measurement audit and a more specific
testable hypothesis; evidence for an operational survey tool remains absent.

## Endpoint and population

The source population is the 922 eligible intervals from the
[recruitment study](recruitment.qmd#source-and-cohort), not all Michigan forest
conditions. The outcome remains a new live sugar-maple sapling tally at
follow-up, using the existing species, diameter, reconciliation, predecessor,
and shared-sampling checks. No eligibility rule is relaxed to obtain more
RI records or events.

Baseline RI is a potential predictor. Follow-up RI is a different measurement
and is not needed to establish baseline length structure when the follow-up
TREE endpoint is otherwise valid. Requiring RI at both visits would answer a
narrower question and could discard valid baseline observations. Neither
condition intervals nor sampled microplots are whole management stands, and
these results are not statewide, survey-weighted population estimates.

## Sampling and protocol definitions

The database dictionary retires `REGEN_SUBP_STATUS_CD` from inventory year
2015 and identifies `REGEN_MICR_STATUS_CD` as the status field populated across
years. Code 1 identifies sampled RI on accessible forest; code 2 is sampled
but without accessible forest on the microplot. Codes 3-5 are nonsampled and
code 9 is historically ambiguous. A blank retired field is therefore not a
reason to reject an otherwise valid code-1 forest sample. Conversely, code 2
cannot verify RI coverage of a positive forest-condition area.
[@burrill2025fiadb, sec. 6.2.10 and 6.2.22]

The baseline screen requires RI plot presence, code-1 coverage of every
contributing condition-microplot slice, consistent subplot identifiers,
compatible core sampling, and a reconstructed condition area matching the
existing microplot proportion. RI seedling records must have valid condition
and subplot-condition foreign keys, species/source/length codes, and counts.
Duplicate keys stop the calculation; invalid records make the affected plot's
RI measurements unavailable. Missing species tallies are not zero-filled until
the measurement frame passes.

The inspected annual supplements specify a May-September RI window and
established seedlings at least two inches long, below one inch DBH. Length is
measured along the seedling from base to terminal bud, not necessarily vertical
height. Counts above five may be estimated. Stump sprouts and suckers from one
origin are not independent stems; only the tallest qualifying stump sprout is
tallied. These details limit biological interpretations of the count totals.
[@nrsri2016, sec. 6.0 and 6.4; @nrsri2018, pp. 18-22]

| RI class | Recorded length | Role in this audit |
|---|---|---|
| 1 | 2 to less than 6 inches | Below the ordinary hardwood minimum |
| 2 | 6 to less than 12 inches | Below the ordinary hardwood minimum |
| 3 | 1 to less than 3 feet | Ordinary-size comparison |
| 4 | 3 to less than 5 feet | Ordinary-size comparison |
| 5 | 5 to less than 10 feet | Prespecified taller group |
| 6 | At least 10 feet | Prespecified taller group |

The codes agree in the inspected 2016-2018 supplements and the v9.4
dictionary. Source code 1 denotes other seedlings and code 2 stump sprouts;
code 3 is reserved for selected large-seeded species, not sugar maple.
[@nrsri2017, pp. 21-22; @burrill2025fiadb, sec. 6.3.14-16]

The inspected supplements and the FIA dictionary define the length schema used
here. They provide the documentary basis for this audit, but they do not
substitute for a year-by-year review of every field supplement.
[@mcwilliams2015ri, pp. 18-23 and 72]

### Historical coverage and date limitations

The full relevant sections of the 2016, 2017, and 2018 annual supplements were
inspected. These years cover 37 available baseline intervals with four events.
The 53 older intervals have dictionary and historical technical-report support,
but their individual 2012-2015 annual supplements were not recovered. Full
version-by-version protocol equivalence is not certified.

The official archive's 2016 copy is marked DRAFT. Its length boundaries agree
with the inspected 2017/2018 supplements and dictionary, but a final deployed
2016 edition is not certified. Older generic cross-references also retain
inconsistent subplot/status wording; the audit uses the explicit microplot
status table and dictionary. These qualifications are retained in the
[data dictionary](../docs/data-dictionary.md) and [analysis decisions](../docs/analysis-decisions.md).

`PLOT.MANUAL` is not treated as a precise P2+ supplement edition: for example,
baseline records store 7.0 while the available annual supplements include 7.1
and 7.2. Year-specific documentary coverage is recorded separately. Plot
measurement dates provide a plausibility screen, not proof of the exact RI
measurement day or leaf condition; the RI tables expose no separate field date.

## Eligibility and valid zeros

```{r}
#| label: tbl-ri-flow
#| tbl-cap: "Baseline RI eligibility and narrower sensitivity populations. Rows after the primary available cohort are separate restrictions, not a single sequential funnel."
flow |> dplyr::transmute(Screen = .data$screen, Conditions = .data$conditions,
  Plots = .data$physical_plots, Counties = .data$counties, `Recruit-bearing` = .data$events) |>
  knitr::kable(align = c("l", "r", "r", "r", "r"))
```

All 90 baseline records pass the implemented frame, record, and date checks.
They cover 311 contributing microplot-condition intervals and have exact mapped
overlap in the existing recruitment frame. This is a qualified local
eligibility finding, not independent certification of all field measurements.
The other 832 eligible recruitment intervals have no baseline RI plot record;
their length-class values remain unavailable, not zero.

At follow-up, two intervals fail the RI frame checks and two more have plot
dates outside the RI window, leaving 86 under a both-visit restriction. These
four flags do not invalidate their baseline RI or the existing TREE endpoint.
The date flags require clarification before using follow-up RI to estimate
length-class changes.

```{r}
#| label: tbl-ri-zeros
#| tbl-cap: "Maple tally states on the 90 baseline RI-available intervals."
zeros |> dplyr::transmute(`Baseline state` = .data$screen, Conditions = .data$conditions,
  Plots = .data$physical_plots, `Recruit-bearing` = .data$events) |> knitr::kable()
```

The 19 conditions without ordinary baseline seedling detections split into nine
with no qualifying RI maple tally and ten containing only smaller RI seedlings.
Thus an ordinary zero is not equivalent to an all-size RI zero. None of those
19 had the defined recorded entry in this snapshot; that does not imply that
small seedlings can never advance or that recruitment cannot follow a zero.

## Count reconciliation

The ordinary maple count equals the sum of RI classes 3-6 in every one of the
311 matched microplot-condition intervals and all 90 condition intervals.
Both totals are 2,658. Including the two smaller RI classes raises the recorded
total to 9,489; four of those tallies use the stump-source category.

```{r}
#| label: tbl-ri-reconciliation
#| tbl-cap: "Same-area comparison of ordinary counts and RI counts at least one foot long."
reconciliation |> dplyr::transmute(Unit = .data$unit, Comparisons = .data$observations,
  `Exact matches` = .data$exact_count_agreements, `Ordinary total` = .data$ordinary_count,
  `RI >=1 ft total` = .data$ri_at_least_one_foot) |> knitr::kable()
```

This establishes numerical redundancy of the collapsed counts in this sample,
not independence of the two measurements. Shared or derived tally information
is a plausible explanation for exact agreement; the database-generation code
has not been audited, so a specific derivation mechanism is not asserted.
It would be misleading to count agreement as independent field validation.

The candidate additional information is the distribution among length classes
and the smaller-size counts. Comparing all 9,489 RI tallies with the 2,658
ordinary tallies as though they measured the same size range would confound
measurement definition with information gain. Neither denominator supports an
individual seedling-to-sapling survival rate.

## Descriptive length support

The at-least-five-foot grouping was fixed before calculating these associations,
based on the published RI schema and the recruitment literature's emphasis on
taller seedlings. Harris and colleagues already showed that length classes can
improve recruitment prediction in a larger regional FIA sample; the present
Michigan result is not a novel demonstration of that general idea.
[@harris2022]

```{r}
#| label: tbl-ri-height
#| tbl-cap: "Recorded recruitment by baseline taller-seedling presence. Intervals are descriptive 95% county-bootstrap intervals, not forecast-validation intervals."
height |> dplyr::transmute(Population = .data$population,
  `Maple >=5 ft` = ifelse(.data$group == "RI maple at least five feet present", "Present", "Not recorded"),
  Conditions = .data$conditions, `Recruit-bearing` = .data$events,
  `Entry fraction` = pct(.data$event_fraction),
  `95% interval` = paste0(pct(.data$lower), " to ", pct(.data$upper))) |> knitr::kable()
```

The primary contrast is `r pct(tall$event_fraction)` versus
`r pct(short$event_fraction)`. Restricting to the 71 ordinary-detected conditions
removes the easiest ordinary-zero group, but still does not adjust for
differences in counts, interval length, site, or existing maple stages. A tall
seedling is already closer in size to the recorded sapling boundary; this
association must not be reinterpreted as a causal effect of height or browsing.

The intervals resample whole counties 2,000 times without fitting a prediction
model. The one-event group is particularly fragile: its bootstrap distribution
can contain no events, so a zero lower bound is not evidence of zero underlying
risk. The intervals omit question selection, historical-protocol uncertainty,
measurement error, and uncertainty about the intended survey decision.

```{r}
#| label: tbl-ri-classes
#| tbl-cap: "Support for the six RI classes. Conditions can occur in several rows; event counts must not be summed across classes."
classes |> dplyr::transmute(Class = .data$length_class, `Recorded tally` = .data$recorded_count,
  `Conditions with class` = .data$conditions_with_class,
  `Recruit-bearing among those` = .data$events_among_conditions_with_class) |> knitr::kable()
```

## Forecasting feasibility and next decision

A development/assessment split following the existing calendar boundary has
53 development intervals with ten events and 37 later intervals with four
events, with no shared physical plots across the split. One missed assessment
event would change observed event recall by 25 percentage points. These already
explored outcomes are not an untouched validation set even if their length
predictors were previously unused.

As a precision illustration only, even a hypothetical rule finding all 14 event
plots would have an exact two-sided binomial 95% lower recall bound of
`r pct(precision$exact_two_sided_95_lower_if_all_found)` under independent-event
sampling. This is not the observed rule's recall, a clustered confidence interval,
or a sample-size justification. Geographic dependence and a held-out sample
would further constrain what can be learned.

The evidence supports retaining length structure as the next measurement
hypothesis. It does not support a six-class fitted forecasting model or an
operational score from this subset. A useful next study would fix a single
supportable length representation, retain ordinary counts and sampling area as
comparators on identical cases, and evaluate the paired gain in a separate,
adequately supported inventory sample. A meaningful gain, forecasting horizon,
decision target, action unit, and measurement cost must be specified first.

Within the existing Michigan snapshot, the baseline measurement milestone is
complete with the historical-access qualifications above. The defensible
contribution is the distinction between seedling abundance, length structure,
and recorded sapling entry, together with explicit limits on usefulness.
Acquiring other inventories or coordinating field work requires a separate
scope decision; neither has occurred. The
[survey validation plan](../docs/survey-validation-plan.md) records these gates.

The subsequent [count-versus-length design](length-study.qmd) fixes a one-term
comparison and planning scenarios. Its outcome-free historical opportunity
screen finds only six physical plots beyond the current recruitment cohort
among positive-seedling opportunities, before full eligibility checks. It does
not establish a new independent Michigan test sample.

## Reproducibility

Run `make ri-audit` after the recruitment dataset exists. The
[eligibility table](../outputs/tables/recruitment-ri-eligibility.csv),
[matched-count results](../outputs/tables/recruitment-ri-count-reconciliation.csv),
[length support](../outputs/tables/recruitment-ri-height-support.csv), and
[split support](../outputs/tables/recruitment-ri-split-support.csv) contain the
aggregate evidence. Record-level joins remain local and Git-ignored.

The [data dictionary](../docs/data-dictionary.md),
[analysis decisions](../docs/analysis-decisions.md), and linked results tables
distinguish declared analytical rules, documentary access, local results, and
limitations.

## References

::: {#refs}
:::
